---
title: "Tutorial 11 In-Class Check - Version D"
format:
  html:
    embed-resources: true
---

## Instructions

Answer all five questions.

Select the best answer in each case.

### Question 1

What is the main purpose of a training/test split in this forecasting tutorial?

A. To build predictions on past games and check whether they work on later games not used to create the model.

B. To make the data easier to print.

C. To guarantee that every forecast error equals zero.

D. To remove the business need for staffing buffers.

### Question 2

Which statement best describes a cell-means forecast?

A. It predicts future observations using the average outcome within a relevant training-data group.

B. It predicts attendance using only the row number in the test set.

C. It fits a separate polynomial for every game.

D. It replaces all missing values with the overall test-set mean.

### Question 3

Suppose day/night has only a modest difference in average attendance, while opponent has larger differences across teams.

Why should a student still be cautious about assuming the opponent model will perform best on future games?

A. Large training differences may reflect noisy small groups and may not generalise well.

B. Opponent cannot be observed before the game.

C. Test-set metrics are never needed for opponent models.

D. Opponent models do not use attendance.

### Question 4

Which code correctly calculates absolute error after `error` has been created?

A.

```{r}
#| eval: false
abs_error = abs(error)
```

B.

```{r}
#| eval: false
abs_error = mean(attendance)
```

C.

```{r}
#| eval: false
abs_error = error ^ 2
```

D.

```{r}
#| eval: false
abs_error = predicted_attendance / attendance
```

### Question 5

What does MAPE add to the model comparison?

A. It expresses typical absolute forecast error as a percentage of actual attendance.

B. It identifies which games belong in the training set.

C. It replaces the need for predictions.

D. It counts how many opponents appear in the data.

::: {.callout-tip collapse="true"}
## Solution Key

1. **A**
   The split lets students build models from earlier seasons and evaluate forecasts on later seasons the model did not use.

2. **A**
   A cell-means model uses training averages within groups such as day/night, day of week x day/night, or opponent.

3. **A**
   A feature can look useful in the training data but fail to generalise if detailed groups are based on limited observations.

4. **A**
   Absolute error is the magnitude of the forecast error, ignoring whether the prediction was too high or too low.

5. **A**
   MAPE reports absolute error relative to actual attendance, making errors easier to interpret as percentages.
:::
