---
title: "Tutorial 11 In-Class Check - Version A"
format:
  html:
    embed-resources: true
---

## Instructions

Answer all five questions.

Select the best answer in each case.

### Question 1

The concessions manager needs to roster staff before the previous game has been played.

Why is a lag-1 attendance model less useful for this operational decision?

A. It can only be used when attendance is measured as a percentage.

B. It relies on information from the previous game, which may not yet be available when rosters must be set.

C. It always overfits more than an opponent-based model.

D. It cannot be evaluated with MAE, RMSE, or MAPE.

### Question 2

The tutorial splits the data so games before 2013 are used for training and games from 2013 onward are used for testing.

What is the best reason for using this time-based split?

A. It mirrors using past games to forecast future games.

B. It guarantees the same number of day and night games in each sample.

C. It removes the need to calculate test-set errors.

D. It ensures every opponent appears exactly once.

### Question 3

Consider the code below:

```{r}
#| eval: false
average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )
```

What kind of model is this creating?

A. A cell-means model that predicts using the average attendance for day and night games in the training set.

B. A regression model with a degree-8 polynomial.

C. A random train/test split.

D. A model that uses the test-set average as the prediction.

### Question 4

A student joins the day/night training means to the test set using:

```{r}
#| eval: false
test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = "day_night")
```

What is the purpose of this join?

A. To remove all games that were played at night.

B. To attach the relevant training-set average attendance to each test-set game as a prediction.

C. To calculate RMSE directly.

D. To randomly assign games to training and test samples.

### Question 5

Which forecast accuracy metric is most directly useful for thinking about a staffing buffer because it is measured in fans?

A. MAPE

B. RMSE

C. MAE

D. R-squared

::: {.callout-tip collapse="true"}
## Solution Key

1. **B**
   Lag-1 attendance may be informative, but it depends on the previous game's attendance and may arrive too late for advance roster planning.

2. **A**
   A time-based split fits on earlier seasons and evaluates on later seasons, which matches the forecasting problem.

3. **A**
   The code calculates training-set averages by `day_night`, so future day and night games can be assigned those group averages.

4. **B**
   The join matches each test game to the appropriate training-set day/night average, turning the mean into a prediction.

5. **C**
   MAE is an average absolute error in the original attendance units, so it translates naturally into a staffing buffer.
:::
