---
title: "Tutorial 11 In-Class Check - Version C"
format:
  html:
    embed-resources: true
---

## Instructions

Answer all five questions.

Select the best answer in each case.

### Question 1

Why does the tutorial focus on predictors known before the game?

A. Because the business decision is made before game-day attendance is observed.

B. Because future information always makes models less accurate.

C. Because test-set data cannot contain calendar variables.

D. Because day/night is the only variable in the dataset.

### Question 2

Which code correctly creates the training sample used in the preparation exercise?

A.

```{r}
#| eval: false
bos_2005_2015_fenway |>
  filter(year < 2013)
```

B.

```{r}
#| eval: false
bos_2005_2015_fenway |>
  filter(year >= 2013)
```

C.

```{r}
#| eval: false
bos_2005_2015_fenway |>
  group_by(year < 2013)
```

D.

```{r}
#| eval: false
bos_2005_2015_fenway |>
  summarise(year < 2013)
```

### Question 3

A student calculates average attendance by `dotw` and `day_night` in the training data.

What must they join on when attaching these predictions to the test set?

A. Only `attendance`.

B. Only `year`.

C. Both `dotw` and `day_night`.

D. Both `MAE` and `RMSE`.

### Question 4

In the error table, the tutorial uses:

```{r}
#| eval: false
error = attendance - predicted_attendance
```

What does this value represent?

A. The amount by which actual attendance differs from the model prediction for a game.

B. The number of games in each group.

C. The average attendance in the training set.

D. The percentage of games played at night.

### Question 5

Which recommendation best reflects the tutorial's model-choice logic for a staffing manager?

A. Always choose the most granular model, because more groups always generalise better.

B. Choose a model by considering test-set accuracy, timing, explainability, and whether metric differences matter operationally.

C. Choose the model with the highest training-set average attendance.

D. Ignore MAE if RMSE has already been calculated.

::: {.callout-tip collapse="true"}
## Solution Key

1. **A**
   Rosters are set before realised attendance is known, so the inputs must be available in advance.

2. **A**
   The training sample uses earlier seasons, so the correct filter is `year < 2013`.

3. **C**
   A two-way cell-means model needs both keys; joining on only one would attach the wrong averages.

4. **A**
   Forecast error is actual attendance minus predicted attendance for each test-set game.

5. **B**
   The tutorial frames model choice as an operational decision, not just a search for the most complex model or one metric.
:::
