<!-- question-type: inclass -->
### Exercise 3: Fitting regression models

Exercises 1 and 2 used simple cell-means models: each forecast came from the average attendance for one group of games. We now turn to ordinary least squares (OLS) regression, which among other things lets us combine multiple predictors in a single model.

The aim is not to find the most complicated model possible. The aim is to ask whether a richer model improves the forecast enough to justify the extra complexity for staffing decisions.

::: {.callout-note title="What is OLS regression?" collapse="true"}
Ordinary least squares (OLS) regression is a way to model the relationship between an outcome we want to predict and one or more predictors. Here, the outcome is attendance, and the predictors might include opponent, game time, day of week, or game number.

OLS works by choosing the line, curve, or set of group differences that makes the model's prediction errors as small as possible in the training data. More precisely, it chooses the coefficients that minimise the sum of squared errors: the squared differences between the actual attendance values and the model's fitted values.

For example, a simple regression might estimate how attendance changes between day and night games. A richer regression can combine several predictors at once, such as opponent, day/night, and day of week. When a predictor is categorical, such as opponent, OLS compares each category to a reference category.

In this tutorial, we use OLS as a forecasting tool. We fit the model on earlier seasons, use it to predict attendance in later seasons, and then evaluate whether those predictions are accurate enough and explainable enough to support staffing decisions.
:::

**(a).** Fit an OLS model that uses the same forward-available predictors we studied earlier: opponent, day/night, and day of week. Complete and run the code below.

```{r}
#| eval: false

basic_ols_model <- 
  lm(YOUR_CODE_HERE ~ YOUR_CODE_HERE + YOUR_CODE_HERE + YOUR_CODE_HERE, 
     data = YOUR_CODE_HERE                                               
  )
```

**(b).** Run the code below to view the model summary.

```{r}
#| eval: false

summary(basic_ols_model)
```

Discuss with your group:

- Which predictors appear to have the largest associations with attendance?
- What does the R-squared value tell us about in-sample fit?
- Why should we avoid choosing the model on R-squared alone?

**(c).** The basic OLS model combines categorical predictors. We can also ask whether attendance changes systematically over the course of the season. Run the two plots below to compare a simple quadratic curve with a more flexible degree-8 curve.

```{r}
#| eval: false

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Quadratic Fit of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x ^ 2) + I(x ^ 3) + I(x ^ 4) + 
      I(x ^ 5) + I(x ^ 6) + I(x ^ 7) + I(x ^ 8),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Degree-8 Polynomial of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

Which curve looks more likely to generalise to future games, and why?

**(d).** Let's formally model the plots above. Run the code below to fit two additional OLS specifications: a simple quadratic model using game number, and a more flexible specification that combines opponent, day/night, day of week, and a degree-8 polynomial in game number.

```{r}
#| eval: false

non_linear_ols_model <- 
  lm(attendance ~ gm + I(gm^2), 
     data = training_sample_2005_2012
  )

eighth_poly_ols_model <- 
  lm(attendance ~ opp + day_night + dotw + gm + I(gm^2) + 
     I(gm^3) + I(gm^4) + I(gm^5) + I(gm^6) + I(gm^7) + I(gm^8),
     data = training_sample_2005_2012
  )
```

**(e).** Complete and run the code below to compare in-sample R-squared across the three OLS specifications.

```{r}
#| eval: false

ols_fit_summary <- tibble(
  model = c("Basic OLS", "Quadratic game number", "Degree-8 polynomial OLS"),
  r_squared = c(
    summary(YOUR_CODE_HERE)$r.squared,
    summary(YOUR_CODE_HERE)$r.squared,
    summary(YOUR_CODE_HERE)$r.squared
  )
)

ols_fit_summary
```

Which model fits the training data best? Why is this not enough to choose the forecasting model? In your answer, consider fit, complexity, and operational usefulness.


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Fit OLS models that extend the forward-available predictors from Exercises 1-2.
- Interpret R-squared as in-sample fit, not forecast quality.
- Explain why the best-fitting training model is not automatically the best forecasting model.

✅   **Core Concepts to Highlight**

- OLS combines predictors that cell-means models handled separately.
- R-squared measures training fit, not future forecast accuracy.
- Highly flexible models can fit noise, especially near the edges of the season.
- Operational usefulness matters: a slightly richer model may not be worth using if it is harder to explain.

💬 **Suggested In-Class Prompts** (if needed)

- "Which predictors would a staffing manager know before the game?"
- "Where might the degree-8 curve be fitting noise rather than structure?"
- "What do we still need to know before choosing the model for staffing?"

📌 **Common Misunderstandings**

- Treating R-squared as a test-set accuracy metric.
- Reading factor coefficients without remembering the reference category.
- Filling the R-squared table with model names as text rather than fitted model objects inside `summary()`.
- Assuming every wiggle in a polynomial curve reflects a real seasonal pattern.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Fitting an OLS model with multiple predictors**

```{r}
basic_ols_model <- 
  lm(attendance ~ opp + day_night + dotw,
     data = training_sample_2005_2012
  )
```

**(b). Viewing the model summary**

```{r}
summary(basic_ols_model)
```

Some opponents and game-time categories are associated with attendance, but R-squared only summarises training fit. It does not tell us whether the model will forecast later seasons well.

**(c). Comparing seasonal curves**

```{r}
ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Quadratic Fit of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x ^ 2) + I(x ^ 3) + I(x ^ 4) + 
      I(x ^ 5) + I(x ^ 6) + I(x ^ 7) + I(x ^ 8),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Degree-8 Polynomial of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

The quadratic curve is more likely to generalise because it imposes a simpler seasonal pattern. The degree-8 curve may be fitting random ups and downs in the training data.

**(d). Fitting the polynomial models**

```{r}
non_linear_ols_model <- 
  lm(attendance ~ gm + I(gm^2), 
     data = training_sample_2005_2012
  )

eighth_poly_ols_model <- 
  lm(attendance ~ opp + day_night + dotw + gm + I(gm^2) + 
     I(gm^3) + I(gm^4) + I(gm^5) + I(gm^6) + I(gm^7) + I(gm^8),
     data = training_sample_2005_2012
  )
```

**(e). Comparing in-sample R-squared**

```{r}
ols_fit_summary <- tibble(
  model = c("Basic OLS", "Quadratic game number", "Degree-8 polynomial OLS"),
  r_squared = c(
    summary(basic_ols_model)$r.squared,
    summary(non_linear_ols_model)$r.squared,
    summary(eighth_poly_ols_model)$r.squared
  )
)

ols_fit_summary
```

Use `basic_ols_model`, `non_linear_ols_model`, and `eighth_poly_ols_model` inside `summary()`.

The degree-8 model should have the highest R-squared because it is most flexible. That does not make it the best forecasting model: simpler models may generalise better and may be easier to explain for staffing decisions.

:::

:::

:::
<!-- END PROFILE:r-solutions -->
