<!-- question-type: prepare -->
### Exercise 1: Building advance predictors from the training set

In the lecture, the lag-1 model performed well because recent attendance is informative. But it is not always useful for workforce planning: concession managers often need to roster staff before the previous game has been played. In this tutorial we therefore focus on predictors that are known before the game:

- whether the game is during the day or evening
- the day of the week
- the opponent

Our goal is to use the training data to decide whether these forward-looking features contain useful predictive signal, then turn them into simple test-set forecasts.

::: {.callout-note title="Why split into training and test data?" collapse="true"}
**Overfitting** happens when a model learns patterns that are too specific to the data used to build it. In practice, this means the model is fitting idiosyncratic variation in the training data rather than meaningful structure. Idiosyncratic variation is not systematic, so we should not expect it to reoccur in future games in the same way. The model may therefore look accurate on familiar data, but perform poorly when asked to predict new games.

A training/test split helps us check for this. We build the model using the **training set**, then evaluate it on a separate **test set** that was not used to calculate the model. If the model works well on the test set, we have better evidence that it has learned a useful pattern rather than memorised noise.

Because this is a forecasting problem, we split by time: earlier seasons are used for training, and later seasons are held out for testing. This mirrors how managers would actually use the model: learn from the past, then forecast future games.
:::

**(a).** Create a time-based training and test split. Use games from seasons before 2013 as the training set, and games from 2013 onward as the test set.

```{r}
#| eval: false

training_sample_2005_2012 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  

test_sample_2013_2015 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  
```

Why is a time-based split more realistic than randomly splitting games for this forecasting problem?

**(b).** Start with the simplest forward-looking feature: whether the game is played during the day or at night. Complete and run the code below, then describe the size and direction of the difference in attendance across day vs night games.

```{r}
#| eval: false

training_sample_2005_2012 |>
  YOUR_CODE_HERE(day_night) |>
  YOUR_CODE_HERE(
    ave_attendance = mean(attendance, na.rm = TRUE), 
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = day_night, y = ave_attendance, fill = day_night)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = round(ave_attendance, 0)),
    vjust = -0.5, size = 4, color = "black"
  ) +
  labs(x = "Game Time", y = "Average Attendance", fill = "Game Time") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ggokabeito::scale_fill_okabe_ito() +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**(c).** Now inspect two richer versions of the same idea: attendance across the different day of week-game time combinations, and attendance across the different opponents. Run the code below.

```{r}
#| eval: false

training_sample_2005_2012 |>
  mutate(
    dotw = factor(
      dotw,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  ) |>
  group_by(dotw, day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = dotw, y = ave_attendance, fill = day_night)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  ggokabeito::scale_fill_okabe_ito(name = "Game Time") +
  labs(x = "Day of the Week", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(desc(ave_attendance)) |>
  mutate(opp = factor(opp, levels = opp)) |>
  ggplot(aes(x = opp, y = ave_attendance)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(x = "Opponent", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1)
  )
```

Briefly answer the following:

- Which feature appears most useful for forecasting attendance?
- Which feature would be easiest to explain to a concessions manager?
- Which feature is most likely to be noisy because it creates small groups?

**(d).** Build three cell-means models from the training set, then join those training means to the test set as predictions. 

::: {.callout-note title="What is a cell-means model?" collapse="true"}
A cell-means model predicts using the average outcome within a group, or "cell", in the training data. For example, a day/night cell-means model calculates the average training-set attendance for day games and the average training-set attendance for night games. It then uses those averages as predictions for future day and night games.

This kind of model is simple and explainable: "games like this usually draw about this many fans." The tradeoff is that more detailed (i.e., narrower) cells can become noisy.
:::

Complete the missing pieces in the code below to fit these cell-means models.

```{r}
#| eval: false

average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE, YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_daynight = ave_attendance) |>
  left_join(average_attendance_by_dotw_dn, by = c(YOUR_CODE_HERE, YOUR_CODE_HERE)) |>
  rename(predicted_attendance_dotw_dn = ave_attendance) |>
  left_join(average_attendance_by_opp, by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_opp = ave_attendance)
```

**(e).** Each model uses information available before the game. Before calculating formal error metrics, which one do you expect to perform best on the test set, and why? In your answer, distinguish between a feature that looks useful in the training data and a model that is likely to generalise well to future games.


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Explain why the tutorial moves from lecture time-series baselines to advance predictors.
- Practice a time-based train/test split to prevent look-ahead bias.
- Use training-set cell means to build simple, explainable forecasts for future games.
- Compare feature usefulness, explainability, and sparsity/noise before seeing test metrics.

✅   **Core Concepts to Highlight**

- Lecture bridge: lag-1 can be accurate but too late for advance staffing decisions.
- Forward-available predictors: day/night, day of week, and opponent are known before rosters are built.
- Temporal validation: split by season rather than randomly mixing earlier and later games.
- Cell-means predictors = fit means on train, assign to test by key(s).
- Simplicity vs granularity: more groups may capture more signal, but small groups can create noisier means.

💬 **Suggested In-Class Prompts** (if needed)

- "What information would the concessions manager know one week before the game?"
- "Would you prefer a slightly less accurate model if it were easier to explain and use?"
- "Why might opponent averages look strong in the training data but generalise poorly?"

📌 **Common Misunderstandings**

- Computing means on the test data, which leaks future information.
- Treating the training plots as proof of test-set accuracy.
- Assuming the most detailed grouping must be best.
- Joining the dotw x day/night model with only one key.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Implement a time-based split**

```{r}
training_sample_2005_2012 <-
  bos_2005_2015_fenway |>
  filter(year < 2013)

test_sample_2013_2015 <-
  bos_2005_2015_fenway |>
  filter(year >= 2013)
```

A time-based split is more realistic because we fit using earlier seasons and evaluate on later seasons, which mirrors how a forecast would be used in practice. Random splitting would allow information from later seasons to influence model development for earlier seasons and can make performance look too optimistic.

**(b). Day/night as an advance predictor**

```{r}
training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = day_night, y = ave_attendance, fill = day_night)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = round(ave_attendance, 0)),
    vjust = -0.5, size = 4, color = "black"
  ) +
  labs(x = "Game Time", y = "Average Attendance", fill = "Game Time") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ggokabeito::scale_fill_okabe_ito() +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

Night games have higher average attendance than day games in the training data. The difference is not enormous, but it is operationally useful because game time is known well before staffing decisions need to be made.

**(c). Comparing richer feature groupings**

```{r}
training_sample_2005_2012 |>
  mutate(
    dotw = factor(
      dotw,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  ) |>
  group_by(dotw, day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = dotw, y = ave_attendance, fill = day_night)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  ggokabeito::scale_fill_okabe_ito(name = "Game Time") +
  labs(x = "Day of the Week", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(desc(ave_attendance)) |>
  mutate(opp = factor(opp, levels = opp)) |>
  ggplot(aes(x = opp, y = ave_attendance)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(x = "Opponent", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1)
  )
```

Day/night is easiest to explain. Day of week x day/night adds useful operational context but creates more groups. Opponent can show large differences, especially for rivalry or less common opponents, but it is also the feature most exposed to sparse groups and noisy averages.

**(d). Build and join the three cell-means predictors**

```{r}
average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(dotw, day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = "day_night") |>
  rename(predicted_attendance_daynight = ave_attendance) |>
  left_join(average_attendance_by_dotw_dn, by = c("dotw", "day_night")) |>
  rename(predicted_attendance_dotw_dn = ave_attendance) |>
  left_join(average_attendance_by_opp, by = "opp") |>
  rename(predicted_attendance_opp = ave_attendance)
```

**(e). Expected test-set performance**

Before evaluating the test set, a reasonable expectation is that day of week x day/night may improve on day/night alone by adding scheduling context. But it may not improve much if the extra groups are noisy. Opponent may look useful in the training data because some opponents have noticeably higher or lower average attendance, but it is the riskiest simple model because some opponent averages are based on relatively few games.

The key distinction is between apparent training-set signal and likely future performance. A feature can look useful in the training data because of real structure, random variation, or both. The test-set metrics in Exercise 2 will help us judge whether the model has captured a pattern that generalises to later seasons.

:::

:::

:::
<!-- END PROFILE:r-solutions -->
