<!-- question-type: prepare -->
### Exercise 2: Evaluating advance predictors in the test set

The lecture introduced MAE, RMSE, and MAPE. Rather than rebuild the same error pipeline three separate times, we will use one tidy workflow to compare all three advance-predictor models from Exercise 1.

**(a).** Run the code below to put the three prediction columns into a long format. Before doing so complete the missing lines that calculate forecast errors.

```{r}
#| eval: false

simple_model_errors <- 
  test_sample_2013_2015 |>
  select(
    attendance,
    predicted_attendance_daynight,
    predicted_attendance_dotw_dn,
    predicted_attendance_opp
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_daynight = "Day/Night",
      predicted_attendance_dotw_dn = "Day of week x Day/Night",
      predicted_attendance_opp = "Opponent"
    ),
    error            = YOUR_CODE_HERE - YOUR_CODE_HERE,
    abs_error        = abs(YOUR_CODE_HERE),
    squared_error    = YOUR_CODE_HERE ^ 2,
    percentage_error = (YOUR_CODE_HERE / YOUR_CODE_HERE) * 100
  )
```

What does `pivot_longer()` do here, and why is it helpful when we want to compare several models?

**(b).** Complete and run the code below to calculate MAE, RMSE, and MAPE for each model.

```{r}
#| eval: false

simple_model_metrics <- 
  simple_model_errors |>
  group_by(model) |>
  summarise(
    MAE  = mean(YOUR_CODE_HERE, na.rm = TRUE),
    RMSE = sqrt(mean(YOUR_CODE_HERE, na.rm = TRUE)),
    MAPE = mean(abs(YOUR_CODE_HERE), na.rm = TRUE),
    .groups = "drop"
  )
```

Which metric is most directly useful for setting a staffing buffer? Which metric is most sensitive to occasional large misses?

**(c).** Run the code below to plot the metrics.

```{r}
#| eval: false

simple_metrics_long <-
  simple_model_metrics |>
  pivot_longer(
    cols = c(MAE, RMSE, MAPE),
    names_to = "metric",
    values_to = "value"
  )

ggplot(simple_metrics_long, aes(x = model, y = value, fill = model)) +
  geom_col() +
  geom_text(aes(label = round(value, 0)), vjust = -0.5, size = 3) +
  facet_wrap(~ metric, scales = "free_y") +
  ggokabeito::scale_fill_okabe_ito() +
  labs(
    title = "Forecast Accuracy Metrics by Model",
    x = "Model",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**(d).** Based on the table and plot, choose one model for a concessions staffing manager. In your answer, consider:

- test-set accuracy
- whether the predictor is available early enough for rostering
- whether the model is simple enough to explain
- whether the accuracy differences are large enough to matter operationally


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Apply MAE, RMSE, and MAPE from the lecture to forward-looking tutorial predictors.
- Use one tidy comparison workflow rather than repeated model-specific error code.
- Interpret test-set performance in light of operational usefulness, not just numerical ranking.
- Connect forecast error magnitudes to staffing buffers and service-risk tradeoffs.

✅   **Core Concepts to Highlight**

- The lecture already introduced the metric formulas; the tutorial applies them to models that can be used before the game.
- Long format creates one row per model prediction, so one grouped summary compares every model.
- MAE is the most intuitive staffing-buffer metric; RMSE highlights models with occasional large misses.
- Small metric differences may not justify a more complex or less explainable model.
- The operational question is not "Which model wins by one decimal point?" but "Which model is accurate enough, early enough, and explainable enough?"

💬 **Suggested In-Class Prompts** (if needed)

- "If MAE is about 2,500 fans, how might that affect the staffing buffer?"
- "Would you choose a model with slightly worse MAE if managers trusted it more?"
- "What would make an opponent-based model risky for future seasons?"

📌 **Common Misunderstandings**

- Treating percentage error and absolute error as interchangeable.
- Thinking the lowest metric always determines the business decision.
- Forgetting that all predictions must be formed from training-set means.
- Over-reading small differences between models that are operationally similar.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Build one long error table**

```{r}
simple_model_errors <- 
  test_sample_2013_2015 |>
  select(
    attendance,
    predicted_attendance_daynight,
    predicted_attendance_dotw_dn,
    predicted_attendance_opp
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_daynight = "Day/Night",
      predicted_attendance_dotw_dn = "Day of week x Day/Night",
      predicted_attendance_opp = "Opponent"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  )
```

`pivot_longer()` stacks the three prediction columns into one `predicted_attendance` column and records the model name in a separate `model` column. This lets us calculate errors and metrics once, then compare models with `group_by(model)`.

**(b). Summarise model accuracy**

```{r}
simple_model_metrics <- 
  simple_model_errors |>
  group_by(model) |>
  summarise(
    MAE  = mean(abs_error, na.rm = TRUE),
    RMSE = sqrt(mean(squared_error, na.rm = TRUE)),
    MAPE = mean(abs(percentage_error), na.rm = TRUE),
    .groups = "drop"
  )
```

MAE is most directly useful for staffing buffers because it is measured in fans. RMSE is most sensitive to occasional large misses because it squares errors before averaging them.

**(c). Plot the metrics**

```{r}
simple_metrics_long <-
  simple_model_metrics |>
  pivot_longer(
    cols = c(MAE, RMSE, MAPE),
    names_to = "metric",
    values_to = "value"
  )

ggplot(simple_metrics_long, aes(x = model, y = value, fill = model)) +
  geom_col() +
  geom_text(aes(label = round(value, 0)), vjust = -0.5, size = 3) +
  facet_wrap(~ metric, scales = "free_y") +
  ggokabeito::scale_fill_okabe_ito() +
  labs(
    title = "Forecast Accuracy Metrics by Model",
    x = "Model",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**(d). Model performance and operational recommendation**

On the out-of-sample test set, Day/Night has the lowest MAE and RMSE, with Day of week x Day/Night essentially tied. Opponent is modestly worse. MAPE is about 5% for all three models, so the differences are real but not huge.

For a concessions staffing manager, Day/Night is a strong default recommendation: it is available well in advance, easy to explain, and performs at least as well as the richer alternatives here. Day of week x Day/Night is also defensible if managers want slightly more schedule detail. The opponent model is less attractive because it is more granular and may rely on noisier training averages for opponents that appear infrequently.

The decision should not be based only on the lowest metric. The practical question is whether a model is accurate enough, early enough, and explainable enough to support roster planning.

:::

:::

:::
<!-- END PROFILE:r-solutions -->
