<!-- question-type: inclass -->
### Exercise 4: Evaluating regression models

Now that we have fitted our OLS models, we need to test whether the extra complexity actually improves forecasts for future games. This mirrors Exercise 2: train on earlier seasons, evaluate on later seasons, and choose based on usefulness for staffing rather than in-sample fit alone.

**(a).** Complete and run the setup code below. It creates a test sample that contains only opponent levels observed in the training data, then generates test-set predictions from the three OLS models.

```{r}
#| eval: false

known_opp_levels <- 
  training_sample_2005_2012 |> 
  pull(opp) |> 
  unique()

safe_test_sample <- 
  test_sample_2013_2015 |>
  filter(opp %in% known_opp_levels)

safe_test_sample <-
  safe_test_sample |>
  mutate(
    predicted_attendance_basic_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample),
    predicted_attendance_non_linear_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample),
    predicted_attendance_eighth_poly_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample)
  )
```

Why do we need to handle opponent levels carefully before using `predict()` on the test set? What would happen if we didn't first trim our test sample in this way?

**(b).** Run the code below, then inspect the resulting `ols_model_errors` data frame.

```{r}
#| eval: false

ols_model_errors <- 
  safe_test_sample |>
  select(
    attendance,
    predicted_attendance_basic_OLS,
    predicted_attendance_non_linear_OLS,
    predicted_attendance_eighth_poly_OLS
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_basic_OLS = "Basic OLS",
      predicted_attendance_non_linear_OLS = "Quadratic game number",
      predicted_attendance_eighth_poly_OLS = "Degree-8 polynomial OLS"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  )
```

What does this code do? Why is each row now a model-game combination rather than just a game? Why do we create `abs_error`, `squared_error`, and `percentage_error`?

**(c).** Now use `ols_model_errors` to create `ols_model_metrics`, a table like the one below. Your output should show the values for each cell - e.g., each model's MAE, RMSE, etc.

| model | MAE | RMSE | MAPE |
|---|---:|---:|---:|
| Quadratic game number |  |  |  |
| Degree-8 polynomial OLS |  |  |  |
| Basic OLS |  |  |  |

Write the code to produce this table:

```{r}
#| eval: false

# Write your answer here

```

**(d).** Interpret the table from part (c).

Which OLS model performs best out of sample? Why doesn't this match the ranking by training-set R-squared from Exercise 3?


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Evaluate OLS forecasts on the held-out 2013-2015 test period, using only rows compatible with the fitted OLS models.
- Explain the intermediate error table before summarising metrics.
- Build the MAE/RMSE/MAPE table from row-level errors.
- Compare OLS test performance with training R-squared.

✅   **Core Concepts to Highlight**

- OLS prediction requires test factor levels seen in training.
- Test errors, not R-squared, determine the best OLS forecaster.
- MAE, RMSE, and MAPE summarise row-level forecast errors.
- The target table helps students work backwards to `group_by(model)` and `summarise()`.

💬 **Suggested In-Class Prompts** (if needed)

- "Did the model with the best R-squared also have the best test MAE?"
- "Which metric would be most useful for staffing buffers?"
- "Which row-level error column is used to calculate each metric?"
- "What grouping is needed to get one row per model?"

📌 **Common Misunderstandings**

- Assuming more variables always means better forecasts.
- Choosing the highest training R-squared model even when it has worse test-set errors.
- Mixing up `error`, `abs_error`, `squared_error`, and `percentage_error` when calculating metrics.
- Producing one overall metric row instead of one row per model by forgetting `group_by(model)`.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Generate a safe test sample and OLS predictions**

```{r}
known_opp_levels <- 
  training_sample_2005_2012 |> 
  pull(opp) |> 
  unique()

safe_test_sample <- 
  test_sample_2013_2015 |>
  filter(opp %in% known_opp_levels)

safe_test_sample <-
  safe_test_sample |>
  mutate(
    predicted_attendance_basic_OLS =
      predict(basic_ols_model, newdata = safe_test_sample),
    predicted_attendance_non_linear_OLS =
      predict(non_linear_ols_model, newdata = safe_test_sample),
    predicted_attendance_eighth_poly_OLS =
      predict(eighth_poly_ols_model, newdata = safe_test_sample)
  )
```

OLS cannot predict an unseen opponent category. We therefore keep only test rows with opponents observed in training.

**(b). Calculate OLS test-set errors**

```{r}
ols_model_errors <- 
  safe_test_sample |>
  select(
    attendance,
    predicted_attendance_basic_OLS,
    predicted_attendance_non_linear_OLS,
    predicted_attendance_eighth_poly_OLS
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_basic_OLS = "Basic OLS",
      predicted_attendance_non_linear_OLS = "Quadratic game number",
      predicted_attendance_eighth_poly_OLS = "Degree-8 polynomial OLS"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  )
```

The code reshapes the data to one row per game-model combination. The error columns feed the metrics: `abs_error` for MAE, `squared_error` for RMSE, and `percentage_error` for MAPE.

**(c). Calculate OLS test-set metrics**

```{r}

ols_model_metrics <- 
  ols_model_errors |>
  group_by(model) |>
  summarise(
    MAE  = mean(abs_error, na.rm = TRUE),
    RMSE = sqrt(mean(squared_error, na.rm = TRUE)),
    MAPE = mean(abs(percentage_error), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(MAE)

```

**(d). Interpret OLS test-set metrics**

The quadratic game-number model has the lowest test MAE. This differs from the training R-squared ranking because higher in-sample fit can reflect overfitting rather than better forecasting.

:::

:::

:::
<!-- END PROFILE:r-solutions -->
