<!-- question-type: prepare -->
### Exercise 1: Exploring and predicting from the training set

Past attendance can tell us a great deal about future attendance, but so too can other factors. To begin, we are going to explore how the following relate to attendance:

- Whether a game takes place during the day or evening
- On which day of the week a game takes place
- Against whom the Red Sox compete in a game

If these factors explain variation in attendance, then we can use them to predict attendance at future games.

**(a).** Ultimately, we want to fit and evaluate our prediction models on different data sets to avoid overfitting. As such, we should undertake our exploratory analyses on the training set too. 

Complete and run the code below to create a training and test split of the data. Use a time-based split: games from seasons before 2013 should be in the training set, and games from 2013 and later should be in the test set.

```{r}
#| eval: false

training_sample_2005_2012 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  

test_sample_2013_2015 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  
```

**(b).** Complete and run the code below to produce a bar chart that shows how mean attendance varies across games played during the day and those played during the evening. What does this plot reveal?

```{r}
#| eval: false

training_sample_2005_2012 |>
  YOUR_CODE_HERE(day_night) |>
  YOUR_CODE_HERE(ave_attendance = mean(attendance, na.rm = TRUE), 
                 .groups = "drop") |>
  ggplot(aes(x = day_night, y = ave_attendance, fill = YOUR_CODE_HERE)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = round(ave_attendance, 0)),
    vjust = -0.5, size = 4, color = "black"
  ) +
  labs(x = "Game Time", y = "Average Attendance", fill = "Game Time") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ggokabeito::scale_fill_okabe_ito() +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**(c).** Let's take these means from our training set and use them to predict attendance in the test set. This gives us a simple prediction model that assigns the average attendance for day games in the training sample to all day games in the test set, and similarly for night games.

Complete and run the code below to join these means to our test sample as predicted attendance values.

```{r}
#| eval: false

average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  YOUR_CODE_HERE(average_attendance_by_daynight, by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_daynight = YOUR_CODE_HERE) 
```

**(d).** Next, let's consider a slightly more complex model where attendance varies as a function of whether a game is played during the day or in the evening AND on which day of the week the game takes place. 

Complete and run the code below to produce a bar chart that shows how mean attendance varies as a function of the interaction between day of the week and time of the day. What does this plot reveal?

```{r}
#| eval: false

training_sample_2005_2012 |>
  mutate(dotw = factor(dotw,
                       levels = c(YOUR_CODE_HERE))) |>  
  group_by(YOUR_CODE_HERE, YOUR_CODE_HERE) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE), 
    .groups = "drop"
  ) |>
  ggplot(aes(x = YOUR_CODE_HERE, y = YOUR_CODE_HERE, fill = day_night)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  ggokabeito::scale_fill_okabe_ito(name = "Game Time") +
  labs(x = "Day of the Week", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


**(e).** Again, let's take these means from our training set and use them to predict attendance in the test set. This gives us a slightly more complex prediction model that assigns the average attendance for each category of game in the training sample (e.g., Saturday night game) to the corresponding category of game in the test set.

Complete and run the code below to join these means to our test sample as predicted attendance values.

```{r}
#| eval: false

average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE, YOUR_CODE_HERE) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  YOUR_CODE_HERE(average_attendance_by_dotw_dn, 
                 by = c(YOUR_CODE_HERE, YOUR_CODE_HERE)) |>
  rename(predicted_attendance_dotw_dn = YOUR_CODE_HERE)

```


**(f).** Next, let's consider a third model where attendance varies as a function of whom the Red Sox compete against in a game. 

Complete and run the code below to produce a bar chart that shows how mean attendance varies as a function of the Red Sox opponents in a game. What does this plot reveal?

```{r}
#| eval: false

training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE), 
            .groups = "drop") |>
  arrange(desc(ave_attendance)) |>
  mutate(opp = factor(opp, levels = opp)) |>
  ggplot(aes(x = opp, y = ave_attendance)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(x = "Opponent", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1)
  )
```

**(g).** And for one final time, let's take these means from our training set and use them to predict attendance in the test set. This gives us our third prediction model which assigns the average attendance for each opponent in the training sample to games in the test sample that feature that opponent.

Complete and run the code below to join these means to our test sample as predicted attendance values.

```{r}
#| eval: false

average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_opp, 
            by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_opp = YOUR_CODE_HERE)
```



<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Practice a time-based train/test split to prevent look-ahead bias and overfitting.
- Use cell means to build simple, explainable predictors (Day/Night; DOTW×D/N; Opponent).
- Interpret exploratory plots and connect visible patterns to predictive signal.

✅   **Core Concepts to Highlight**

- Temporal validation: split by season (pre-2013 vs 2013+) rather than random rows.
- Cell-means predictors = “fit means on train, assign to test by key(s)”.
- Categorical interactions (DOTW×D/N) vs single splits (D/N).
- Sparsity risk with many categories (Opponent): some cells have few games → noisier means.
- Tidy pipelines: group_by → summarise → join to test.


💬 **Suggested In-Class Prompts** (if needed)

NA

📌 **Common Misunderstandings**

- Using test data to compute means (leakage).
- Mismatched keys in left_join (e.g., join DOTW×D/N using only one key).
- Assuming bigger or more flexible models always better: Opponent split can be noisier.
- Not coercing DOTW order before plotting (alphabetical vs calendar order).


:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Implement a time-based split**

```{r}
training_sample_2005_2012 <-
  bos_2005_2015_fenway |>
  filter(year < 2013)

test_sample_2013_2015 <-
  bos_2005_2015_fenway |>
  filter(year >= 2013)
```

We’ll fit our models on the training sample, then predict on a test sample that our models haven’t seen—so we can gauge generalization and avoid look-ahead bias

**(b). On average, attendance is higher for night games**

```{r}
training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE), 
            .groups = "drop") |>
  ggplot(aes(x = day_night, y = ave_attendance, fill = day_night)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = round(ave_attendance, 0)),
    vjust = -0.5, size = 4, color = "black"
  ) +
  labs(x = "Game Time", y = "Average Attendance", fill = "Game Time") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ggokabeito::scale_fill_okabe_ito() +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```
We can see that, on average, attendance is higher for night games than for day games. This suggests that game time is associated with attendance levels at games—a key insight for our prediction models.

**(c). Using training means for day/night as predictions in the test set**

```{r}
average_attendance_by_daynight <-
  training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
  )

test_sample_2013_2015 <-
  test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = "day_night") |>
  rename(predicted_attendance_daynight = ave_attendance)
```

**(d). Attendance as a function of DOTW x Day/Night**

```{r}
training_sample_2005_2012 |>
  mutate(
    dotw = factor(
      dotw,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  ) |>
  group_by(dotw, day_night) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE), 
            .groups = "drop") |>
  ggplot(aes(x = dotw, y = ave_attendance, fill = day_night)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  ggokabeito::scale_fill_okabe_ito(name = "Game Time") +
  labs(x = "Day of the Week", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Interesting. There is some variation along these dimensions, but not a great deal. The plot shows that weekend night games (Friday and Saturday) draw the largest crowds, while weekday day games (Monday through Thursday) have the lowest attendance. This suggests that both day of the week and game time jointly co-move with attendance, albeit quite weakly.

**(e). Using training means for day/night by day of the week as predictions in the test set**

```{r}
average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(dotw, day_night) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
            )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_dotw_dn, 
            by = c("day_night", "dotw")) |>
  rename(predicted_attendance_dotw_dn = ave_attendance)
```

**(f). Attendance as a function of opponent**

```{r}
training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE), 
            .groups = "drop") |>
  arrange(desc(ave_attendance)) |>
  mutate(opp = factor(opp, levels = opp)) |>
  ggplot(aes(x = opp, y = ave_attendance)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(x = "Opponent", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1)
  )
```
The plot shows that games against certain opponents (e.g., Marlins, Padres) draw much larger crowds than games against other opponents (e.g., Reds, Pirates). This suggests that opponent identity is a strong predictor of attendance.

**(g). Using training means for opponent as predictions in the test set**

```{r}
average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(ave_attendance = mean(attendance, na.rm = TRUE),
            .groups = "drop"
            )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_opp, 
            by = "opp") |>
  rename(predicted_attendance_opp = ave_attendance)
```


:::

:::

:::
<!-- END PROFILE:r-solutions -->
