<!-- question-type: inclass -->
### Exercise 3: Fitting regression models

We now turn to ordinary least squares regression models that combine multiple predictors to forecast attendance. OLS regression is a powerful and interpretable method that estimates the relationship between a dependent variable (attendance) and one or more independent variables (e.g., day of week, opponent, game time).

**(a).** Let's begin by fitting an OLS model that uses opponent, day/night, and day of week. We will follow the same process as before and fit on the training data. Complete and run the code below to fit this model.

```{r}
#| eval: false

basic_ols_model <- 
  lm(YOUR_CODE_HERE ~ YOUR_CODE_HERE + YOUR_CODE_HERE + YOUR_CODE_HERE, 
     data = YOUR_CODE_HERE                                               
  )

```

**(b).** Run the code below to view this model's basic statistics. What do you notice about the coefficients for the different predictors? What does the R-squared value tell you about the model's fit?

```{r}
#| eval: false

summary(basic_ols_model)
```

**(c).** Alternatively, we can add nonlinearity by including polynomial terms. For instance, we might expect attendance to vary nonlinearly with game number (e.g., early-season games are better attended than late-season games, or vice-versa). 

Run the code below. What does this plot suggest about how attendance varies over the course of the MLB season?

```{r}
#| eval: false

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Quadratic Fit of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

**(d).** Now, let's fit a polynomial regression model that includes a quadratic term for game number. Complete and run the code below to fit this model.

```{r}
#| eval: false

non_linear_ols_model <- 
  lm(YOUR_CODE_HERE ~ YOUR_CODE_HERE + I(YOUR_CODE_HERE^2), 
     data = YOUR_CODE_HERE
  )
```

**(e).** Run the code below to view this model's basic statistics. How do the coefficients for the polynomial terms compare to those in the previous model? What does the R-squared value tell you about this model's fit?

```{r}
#| eval: false

summary(non_linear_ols_model)
```

**(f).** Finally, we can fit a highly flexible polynomial model that includes up to an 8th-degree polynomial in game number, along with opponent, day/night, and day of week. This allows us to capture complex seasonal patterns in attendance.

Run the code below. What does this plot suggest about how attendance varies over the course of the MLB season? Why should we be cautious about using such a high-degree polynomial?

```{r}
#| eval: false

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x ^ 2) + I(x ^ 3) + I(x ^ 4) + 
      I(x ^ 5) + I(x ^ 6) + I(x ^ 7) + I(x ^ 8),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Degree-8 Polynomial of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

**(g).** Now, let's fit this high-degree polynomial regression model that also includes opponent, day/night, and day of week. Complete and run the code below to train this model.

```{r}
#| eval: false

eighth_poly_ols_model <- 
  lm(YOUR_CODE_HERE ~ opp + day_night + dotw + gm + I(gm^2) + 
     I(gm^3) + I(gm^4) + I(gm^5) + I(gm^6) + I(gm^7) + I(gm^8),
     data = YOUR_CODE_HERE
  )
```

**(h).** Run the code below to view this model's basic statistics. What does the R-squared value tell you about this model's fit?

```{r}
#| eval: false

summary(eighth_poly_ols_model)
```


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Fit and interpret OLS with categorical predictors (opponent, day/night, day-of-week).

- Incorporate nonlinearity with polynomial terms in a continuous predictor (game number).

- Contrast in-sample fit (R²) with model complexity and discuss risks of overfitting (e.g., degree-8).

- Read OLS output: reference levels, coefficient interpretation, overall F-test, and limits of R².

✅   **Core Concepts to Highlight**

- R² vs usefulness: Higher R² does not guarantee better generalization; use test-set error for that claim.

- Polynomial terms: `I(gm^k)` adds flexibility; higher degrees → potential overfit and edge wiggles.

- Visualization to hypothesize form (quadratic smoother) before fitting.


💬 **Suggested In-Class Prompts** (if needed)

- “Why might many opponent coefficients be insignificant?”

- “What does R² ≈ 0.165 tell us here?”

- “If the pattern is weak, should we still model it? Why/why not?”

- “What risks do high-degree polynomials pose for out-of-sample performance?”

📌 **Common Misunderstandings**

- Misreading baselines: forgetting that factor coefficients are relative to the reference category.

- R² worship: assuming higher R² = better model; ignoring OOS error and operational relevance.

- Polynomial misuse: adding high-degree terms without checking stability/interpretability; not wrapping in I().

- Leakage risk (general reminder): fitting on training only; all plots that motivate form should still use training for downstream modeling choices.

- Over-interpreting noisy smooths: taking small undulations as “real” effects.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a). Fitting an OLS model with multiple predictors**

```{r}
basic_ols_model <- 
  lm(attendance ~ opp + day_night + dotw,
     data = training_sample_2005_2012
  )
```

**(b). Viewing model statistics**

```{r}
summary(basic_ols_model)
```

The intercept is the baseline (day game, reference opponent/day), and most opponent effects are small and not statistically significant; a few are sizable/negative and significant (e.g., CIN −2247, PIT −2291, TBD −1536), while night games add about +275 attendees (p < .01); day-of-week terms are mostly negligible.

R-squared = 0.165 indicates the model explains a modest share of variation—useful signal but a lot left unexplained—though the overall F-test is highly significant, so the predictors improve fit vs. a mean-only model.

**(c). Plotting attendance vs. game number with a quadratic fit**

```{r}
ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Quadratic Fit of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

Attendance follows a gentle inverted-U: it edges up from early games to a mid-season peak, then drifts down toward season’s end (some reasons for this could be that people take a while to 'get into' the baseball season, or that mid-season coincides with the middle of summer when people have time off to attend games, etc. The effect is modest relative to the overall scatter, so game-to-game variability remains large.

**(d). Fitting an OLS model with a quadratic term**

```{r}
non_linear_ols_model <- 
  lm(attendance ~ gm + I(gm^2), 
     data = training_sample_2005_2012
  )
```

**(e). Viewing model statistics**

```{r}
summary(non_linear_ols_model)
```

The coefficients are small and not significant and so suggest the inverted-U shape is weak. R-squared = 0.005 indicates this model explains very little of the variation in attendance. The categorical model with opponent + day/night + dotw had R-squared = 0.165 and was jointly significant—better in-sample fit—though (as noted earlier) lower in-sample R-squared can still yield better out-of-sample errors when the simpler structure generalizes.

**(f). Plotting attendance vs. game number with an 8th-degree polynomial fit**

```{r}
ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x ^ 2) + I(x ^ 3) + I(x ^ 4) + 
      I(x ^ 5) + I(x ^ 6) + I(x ^ 7) + I(x ^ 8),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Degree-8 Polynomial of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

Attendance wiggles gently around ~37k across the season—small undulations with slight bumps mid-season and end-effects, not large systematic swings.
A degree-8 polynomial can overfit noise, creating spurious wiggles (especially near the edges) that don’t generalize out of sample. Such fits are unstable to small data changes, hard to interpret, and can degrade test accuracy versus simpler forms (e.g., quadratic or cross-validated splines).

**(g). Fitting an OLS model with an 8th-degree polynomial term**

```{r}
eighth_poly_ols_model <- 
  lm(attendance ~ opp + day_night + dotw + gm + I(gm ^ 2) + I(gm ^ 3) + 
     I(gm ^ 4) + I(gm ^ 5) + I(gm ^ 6) + I(gm ^ 7) + I(gm ^ 8),
     data = training_sample_2005_2012
  )
```

**(h). Viewing model statistics**

```{r}
summary(eighth_poly_ols_model)
```

R-squared = 0.177 suggests that the models barely improves over the simpler OLS from earlier, indicating limited additional explanatory power (and potential overfitting). The overall F-test is significant, so these predictors improve on a mean-only model, but most of the variation remains unexplained.


:::

:::

:::
<!-- END PROFILE:r-solutions -->